---
title: "Nome-ERA5 Distance"
output: pdf_document
---

Euclidean distance between data observed in Nome and ERA5 output at nearby points was computed to determine which ERA5 data point has the greatest correspondence with observations at Nome. This distance distance metric was calculated based on observations/output of daily snowfall, snow depth (daily average taken for ERA5 data), minimum 2m air temperature and maximum 2m air temperature. 

```{r setup, echo = FALSE}

# compare data from four different cells around Nome to see if 
#   one is clearly more representative than the others

#-- Functions -----------------------------------------------------------------

# create paths to data files
# all vars in one function now, removing var arg
mk_ncfns <- function() {
  suffix <- c("1979-1982", "1983-1986", "1987-1990", "1991-1994",
              "1995-1998", "1999-2002", "2003-2006", "2007-2010",
              "2011-2014", "2015-2018")
  paste0("../raw_data/ERA5/ERA5_sf_sd_ta_Nome_sector_", suffix, ".nc")
}

# extract data from netcdfs
extract_nc <- function(nc_fn, varid) {
  nc <- nc_open(nc_fn)
  arr <- ncvar_get(nc, varid = varid)
  ts <- ncvar_get(nc, varid = "time")
  nc_close(nc)
  dimnames(arr)[[3]] <- ts
  arr
}

# bind arrays and create data.frame
# two modes: sum (e.g. snowfall), avg (e.g. snow depth), 
make_df <- function(arr_lst, varname, fun = sum, conv = FALSE) {
  arr <- abind(arr_lst, along = 3)
  orig <- ymd_hms("1900-01-01 00:00:0")
  tvec <- hours(dimnames(arr)[[3]]) + orig
  
  # make data frame
  # split into each point and combine
  # add i,j grid cell identifier, based on 3x3 matrix overlaid on
  #   x = lon, y = lat coord system
  ij_df <- expand.grid(1:3, 1:3)
  df <- data.frame(var = unlist(asplit(arr, c(1, 2))),
                   ts = rep(tvec, 9), 
                   # use i,j coords for desired orientation on 
                   #   x,y lon,lat grid system; not based on nc array
                   ij = rep(paste(ij_df[, 2], ij_df[, 1], sep = ","), 
                            each = length(tvec)))
  # summarise data by date
  df <- df %>% 
    mutate(date = date(ts)) %>%
    group_by(date, ij) %>%
    summarise(var = fun(var, na.rm = TRUE))
  
  # convert from Kelvin to Celsius if temp var
  if (conv == TRUE) {
    df <- df %>% mutate(var = var - 273.15)
  } 
  
  # rename variable
  names(df)[3] <- varname
  df
}

# prep daily nome data
prep_daily <- function() {
  # only need these vars
  vars <- c("DATE", 
            "SNOW", "SNOW_ATTRIBUTES", "SNWD", "SNWD_ATTRIBUTES", 
            "TMIN", "TMIN_ATTRIBUTES", "TMAX", "TMAX_ATTRIBUTES")
  # better names
  bnames <- c("date", 
              "sf", "sf_attr", "sd", "sd_attr", 
              "tmin", "tmin_attr", "tmax", "tmax_attr")
  nome <- fread("../raw_data/GHCND/Nome.csv", 
                select = vars, col.names = bnames)
  
  # convert to correct type and units (m and C) and 
  #   subset to matching time frame
  begin <- ymd("1979-01-01")
  end <- ymd("2018-12-31")
  nome[, ':=' (date = ymd(date),
               sf = as.numeric(sf)/1000,
               sd = as.numeric(sd)/1000,
               tmin = as.numeric(tmin)/10,
               tmax = as.numeric(tmax)/10)]
  nome <- nome[date >= begin & date <= end, ]
}

# compute distance between ERA5 i,j time series for all variables
# (and each var separately?)
# args will be df and DT, one for ERA5 (sim) and one for Nome (obs)
compute_dist <- function(sim, obs) {
  # drop unnecessary columns
  # tavg missing in 9933 cases for Nome daily. Ignoring from distance calc
  vars <- c("date", "sf", "sd", "tmin", "tmax")
  obs <- obs[, ..vars]
  # remove NAs and make sure same dates are removed from sim, drop dates
  obs <- na.omit(obs)
  valid_dates <- obs$date
  sim <- sim %>% filter(date %in% valid_dates)
  sim$date <- NULL
  obs[, date := NULL]
  # unique cells to calculate distance over
  cells <- levels(sim$ij)
  # compute distance for each cell
  obs <- as.matrix(obs)
  cell_dist <- function(cell, sim, Y) {
    X <- sim %>% 
      ungroup() %>%
      filter(ij == cell) %>% 
      select(-ij) %>% 
      as.matrix()
    sqrt(sum((X - Y)^2))
  }
  
  sapply(cells, cell_dist, sim, obs)
}

# take vector of dists with "i,j" coordinate names and return
#   df with lon/lat and dists
get_coords <- function(dists) {
  # get coords from an .nc file
  nc <- nc_open("../raw_data/ERA5/ERA5_sf_sd_ta_Nome_sector_1979-1982.nc")
  xc <- nc$dim$longitude$vals
  yc <- nc$dim$latitude$vals
  nc_close(nc)
  # create df
  ijs <- names(dists)
  yci <- as.numeric(substr(ijs, 1, 1))
  xcj <- as.numeric(substr(ijs, 3, 3))
  # add grid cell vertices and colors by rank of similarity
  rh <- 0.25/2
  df <- data.frame(lon = xc[xcj], lat = yc[yci], dist = dists)
  df %>%
    mutate(xleft = lon - rh,
           ybottom = lat - rh,
           xright = lon + rh,
           ytop = lat + rh,
           drank = rank(dist),
           col = terrain.colors(9)[drank])
}

plot_dists <- function() {
  # extract lon and lat from nc file and close
  sf_path <- "../raw_data/ERA5/ERA5_land_sea_mask_Nome_sector.nc"
  sf <- nc_open(sf_path)
  xc <- ncvar_get(sf, varid = "longitude")
  yc <- ncvar_get(sf, varid = "latitude")
  nc_close(sf)
  # Nome coords
  Nlon <- -165.406387
  Nlat <- 64.501114
  Nome_coords <- c(Nlon, Nlat)
  Nome_coords[1] <- 360 - Nome_coords[1] * (-1)
  Nome_coords <- data.frame(x = Nome_coords[1], y = Nome_coords[2])
  # grid coords
  coords <- expand.grid(xc, yc)
  plot(coords, pch = 19, xlab = "Lon", ylab = "Lat")
  rect(dist_df$xleft, dist_df$ybottom, dist_df$xright, dist_df$ytop, 
       col = dist_df$col)
  points(coords, pch = 19)
  points(Nome_coords, col = "red", pch = 19)
  text(dist_df$lon, dist_df$lat - 0.08, round(dist_df$dist))
}

#------------------------------------------------------------------------------

#-- Main ----------------------------------------------------------------------

library(ncdf4)
library(data.table)
library(abind)
library(lubridate)
library(dplyr)
library(ggplot2)
library(knitr)

# vars: snowfall, snow depth, min temp, max temp, avg temp
sf <- lapply(mk_ncfns(), extract_nc, "sf") %>% 
  make_df("sf")
sd <- lapply(mk_ncfns(), extract_nc, "sd") %>% 
  make_df("sd", fun = mean)
# temperature vars
t_arr_lst <- lapply(mk_ncfns(), extract_nc, "t2m")
tmin <- t_arr_lst %>% make_df("tmin", min, TRUE)
tmax <- t_arr_lst %>% make_df("tmax", max, TRUE)
tavg <- t_arr_lst %>% make_df("tavg", mean, TRUE)

# ignoring tavg in dist calc, missing 9933 Nome observations in
#   selected time window 
era5 <- cbind(sf, sd, tmin, tmax) %>% 
  select(date, ij, sf, sd, tmin, tmax)

# save era5 data for other analyses
saveRDS(era5, "../Nome_Mets_aux/data/era5.Rds")

# load daily nome data
nome <- prep_daily()

# compare distance between each grid point 
dists <- compute_dist(era5, nome)

# add coords for grid cells and colors
dist_df <- get_coords(dists)

# save distance results in case needed
# saveRDS(dist_df, "data/dist_results.Rds")

```

Below is a plot displaying results. Darker green represents lower distance between Nome and the ERA5 grid point (greater similarity) and lighter colors represent greater distance. Nome is the red dot, and values are the computed distances. ERA5 data were only downloaded for the 9 points colored.

```{r plot, echo=FALSE, fig.height=5, fig.width=5.25, fig.align="center"}
plot_dists()
```

