---
title: "Nome-ERA5 Extremes Comparison"
output: pdf_document
---

Daily observations of three variables at Nome (snowfall, snow depth, and temperature) were summarized for comparison with ERA5 data from the surrounding 9 grid points to inform choice of grid cell(s) to use in further analyses.  

## Results

```{r setup, echo = FALSE}
# compare extreme events between observed values at Nome and 
#   era5 grid cells

# convert C to F
C_to_F <- function(deg_C) {
  deg_C * (9/5) + 32
}

# load and prep daily data
prep_daily <- function(fn) {
  # Daily data 
  # only need these vars
  vars <- c("DATE", 
            "SNOW", "SNOW_ATTRIBUTES", "SNWD", "SNWD_ATTRIBUTES", 
            "TMIN", "TMIN_ATTRIBUTES", "TMAX", "TMAX_ATTRIBUTES", 
            "TAVG", "TAVG_ATTRIBUTES")
  # better rnames
  bnames <- c("date", 
              "sf", "sf_attr", "sd", "sd_attr", 
              "tmin", "tmin_attr", "tmax", "tmax_attr", 
              "tavg", "tavg_attr")
  nome <- fread(fn, select = vars, col.names = bnames)
  
  # convert to correct type and units (m and C) and 
  #   subset to matching time frame
  begin <- ymd("1979-01-01")
  end <- ymd("2018-12-31")
  nome[, ':=' (date = ymd(date),
               year = year(date),
               decade = year(date) - year(date) %% 10,
               sf = as.numeric(sf)/1000,
               sd = as.numeric(sd)/1000,
               tmin = C_to_F(as.numeric(tmin)/10),
               tmax = C_to_F(as.numeric(tmax)/10),
               tavg = C_to_F(as.numeric(tavg)/10))]
  nome[, ym := format(date, "%Y-%m")]
  nome <- nome[date >= begin & date <= end, ]
}

# sumarize era5 temp data (count number of extremes each year)
sumi_era5_tmin <- function(df, decade = FALSE, thr = -40) {
  df <- df %>% mutate(year = year(date))
  if(decade) {
    df %>% 
      mutate(decade = year(date) - year(date) %% 10,
             tmin = C_to_F(tmin)) %>%
      group_by(decade, ij) %>%
      summarise(count = sum(tmin <= thr))
  } else {
    df %>%
      group_by(year, ij) %>%
      summarise(count = sum(tmin <= thr),
                max_sd = max(sd))
  }
}

# count totals of extremes and counts in common with observed
total_era5_tmin <- function(df, obs_DT) {
  df <- df %>%
    left_join(obs_DT, by = "year") %>%
    mutate(common = pmin(count, obs_count)) %>%
    group_by(ij) %>%
    summarise(total = sum(count), 
              common = sum(common))
  return(df)
}

# summarize era5 snowfall
sumi_era5_sf <- function(df) {
  df %>%
    mutate(ym = format(date, "%Y-%m"),
           month = month(date)) %>%
    group_by(month, ym, ij) %>%
    summarise(sum_sf = sum(sf)) %>%
    group_by(month) %>%
    summarise(avg_sf = mean(sum_sf))
}

# summarize era5 snow depth
sumi_era5_sd <- function(df) {
  df %>%
    mutate(year = year(date)) %>%
    group_by(year, ij) %>%
    summarise(max_sd = max(sd))
}

# summarize Nome temps
sumi_Nome_tmin <- function(DT, decade = FALSE, thr = -40) {
  if(decade) {
    DT <- DT[, keyby = .(decade), .(obs_count = sum(tmin <= thr))]
  } else {
    DT <- DT[, keyby = .(year), .(obs_count = sum(tmin <= thr))] 
  }
  as.data.frame(DT)
}

# summarize Nome snowfall
sumi_Nome_sf <- function(DT) {
  DT <- DT[, keyby = .(month, ym),
           .(sum_sf = sum(sf, na.rm = TRUE))]
  DT <- DT[, keyby = .(month), .(avg_sf = mean(sum_sf))]
  as.data.table(DT)
}

# summarize Nome snow depth
sumi_Nome_sd <- function(DT) {
  DT <- DT[, keyby = .(year),
           .(max_sd = max(sd, na.rm = TRUE))]
  as.data.frame(DT)
}

# compute distance between a vector and each col of df
compute_dist <- function(df, Y) {
  # drop year col
  X <- as.matrix(df[, -1])
  apply(X, 2, function(X, Y) dist(rbind(X, Y)), Y)
}

# make a df of rect boundaries and colors from df of grid center points
#   and land proportions
make_mask_df <- function(df) {
  res <- 0.25/2
  data.frame(xleft = df[, 1] - res, xright = df[, 1] + res,
             ybottom = df[, 2] - res, ytop = df[, 2] + res)
}

# add wgs84 coords to df with ij col
add_coords <- function(df) {
  lon <- c(194.25, 194.5, 194.75)
  lat <- c(64.75, 64.5, 64.25)
  temp <- expand.grid(lon, lat)
  df$lon <- temp[, 1]
  df$lat <- temp[, 2]
  return(df)
}

# make land sea mask df
make_mask <- function(lsm_fn) {
  nc <- nc_open(lsm_fn)
  lon <- ncvar_get(nc, "longitude")
  lat <- ncvar_get(nc, "latitude")
  lsm <- ncvar_get(nc, "lsm")
  nc_close(nc)
  df <- expand.grid(lon, lat)
  names(df) <- c("lon", "lat")
  df$lsm <- round(as.vector(lsm))
  res <- 0.25/2
  df %>%
    mutate(xleft = lon - res,
           ybottom = lat - res,
           xright = lon + res,
           ytop = lat + res)
}

# plot tmin comparison results
plot_tmin <- function(df, mask) {
  # make Nome coords
  Nlon <- -165.406387
  Nlat <- 64.501114
  Nome_coords <- c(Nlon, Nlat)
  Nome_coords[1] <- 360 - Nome_coords[1] * (-1)
  Nome_coords <- data.frame(x = Nome_coords[1], y = Nome_coords[2])
  mask_cols <- c("#3399FF", "#669900")
  plot(mask$lon, mask$lat, xlab = "Lon", ylab = "Lat")
  rect(mask$xleft, mask$ybottom, mask$xright, mask$ytop,
       col = mask_cols[mask$lsm+1])
  text(df$lon, df$lat - 0.05, round(df$total))
  text(df$lon + 0.05, df$lat - 0.055, paste0("(", round(df$common), ")"))
  points(mask$lon, mask$lat, pch = 19)
  points(Nome_coords, col = "red", pch = 19)
}

library(data.table)
library(dplyr)
library(lubridate)
library(ncdf4)
library(knitr)

nome <- prep_daily("data/Nome_daily.csv")

# era 5 data of grid cells surrounding Nome
era5 <- readRDS("data/era5.Rds")

# land sea mask filepath
lsm_fn <- "data/ERA5_land_sea_mask_Nome_sector.nc"
mask <- make_mask(lsm_fn)

# desired outputs

# grid with total counts of temp anomalies (days where tmin <= -40, -35)
# counts of co-ocurring temp anomalies (presented as fraction of total anomalies
#   e.g. 8 anomalies (6/8 co-ocurring in same year))

# count <= tmin threshold for the Nome sector grid cells by year
tmin40_nome <- sumi_Nome_tmin(nome)
tmin35_nome <- sumi_Nome_tmin(nome, thr = -35)
# total nome counts 
tot_40_nome <- tmin40_nome %>% filter(obs_count > 0)
tot_35_nome <- tmin35_nome %>% filter(obs_count > 0)
```

#### Temperature Anomalies

Below are the counts of days where the minimum temperature ($T_{min}$) was $\leq$ -40 degrees (top figure) and -35 degrees (bottom figure). The total count is the number directly below the grid cell center dot, and the number of counts in common with observations at Nome is in parentheses.  

Nome is the red dot.  

```{r plot_tmin, echo = FALSE}

# compare with Nome observations and plot
# -40
sumi_era5_tmin(era5) %>% 
  total_era5_tmin(tmin40_nome) %>% 
  add_coords() %>%
  plot_tmin(mask)
# -35
tmin35_df <- sumi_era5_tmin(era5, thr = -35) %>% 
  total_era5_tmin(tmin40_nome) %>% 
  add_coords() %>%
  plot_tmin(mask)



# grid with distances between max snow depth
#max_sd_era5 <- sumi_era5_sd(era5) %>% tidyr::spread(ij, max_sd)
#max_sd_Nome <- sumi_Nome_sd(nome)
#compute_dist(max_sd_era5, max_sd_Nome$max_sd)

# dotplot with month on x axis and monthly mean values of snowfall

#extr_results <- readRDS("data/extr_results.Rds")


```

Here is a summary of the **observed** number of days in each year where $T_{min}$ fell below -40: `r kable(tot_40_nome)`  

and below -35: `r kable(tot_35_nome)`
